### 2024-03-06 misc/expo/entropy.tex
Done first estimate in PFR

\maketitle{Entropy and additive combinatorics}{}{Marcel K. Goh}{15 February 2024}

\floattext4.5 \ninebf Abstract.
\ninepoint These expository notes give a gentle introduction to the notion of entropy as it is used
in additive combinatorics, with a view towards understanding the proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. Much of the first section has been
transcribed from lectures given by W.~T.~Gowers.

@@

We shall take it on faith that there really exists a function on random variables satisfying
these axioms.
In fact, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.
(It is very possible you have met the formula for entropy
somewhere on your travels, but you will not find it anywhere in
these notes.)

@@

Notationally, we would expect that $\Eta\{Y\given X\} = \Eta\{Y\}$ if $X$ and $Y$ are independent.
This is the first proposition we will carefully prove using the axioms above.

@@

Next we establish the intuitive fact
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

@@

The link between entropy and additive combinatorics is rather a fine one. It is based on the following
observation. (In this and the rest of the notes, $\lg$ denotes the base-$2$ logarithm. This base can
be changed by modifying the normalisation axiom.)

@@

The submodularity inequality is often stated in terms of a triple of random variables in terms of
the {\it conditional mutual information}, which is defined by
$$\II\{ X : Y \given Z\} = \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\}.$$

@@

\proof Since conditioning does not increase entropy, we have
$$\Eta\{ X\pm Y\} \ge \Eta\{ X\pm Y \given Y\},$$
and since the probabilities $\pr\{ X + Y = z \given Y = y\} = \pr\{ X = z-y \given Y = y\}$ for all $z\in G$,
by invariance we have
$$\Eta\{ X\pm Y\} \ge \Eta\{X\given Y\} = \Eta\{X\} - \II\{X : Y\}.$$
Repeating the same argument but exchanging the roles of $X$ and $Y$, we get
$$\Eta\{ X\pm Y\} \ge \Eta\{Y\given X\} = \Eta\{Y\} - \II\{X : Y\},$$
so
$$\Eta\{ X\pm Y\} \ge \max\bigl(\Eta\{X\}, \Eta\{Y\}\bigr) - \II\{X : Y\}.$$
Now let $Z$ be any random variable with finite support.
$$\eqalign{
\Eta\{ X\pm Y\given Z\} &= \sum_{z\in G} \pr\{Z = z\} \Eta\{ X \pm Y \given Z = z\} \cr
&\ge \Bigl(\max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} \Bigr)
\sum_{z\in G} \pr\{Z = z\} \cr
&= \max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} ,\cr
}$$
which completes the proof.\slug

@@

\medskip\boldlabel Entropic Ruzsa distance.
In additive combinatorics, whenever we have two finite subsets $A$ and $B$
of the same abelian group, we can compute the Ruzsa distance
$$d(A,B) = \lg {|A-B|\over \sqrt{|A|\cdot|B|}}$$
between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
$A$.)

The entropic analogue of the Ruzsa distance is defined as follows.
For finitely supported random variables $X$ and $Y$ taking values in the same abelian group,
we let $X'$ and $Y'$ be independent copies of $X$
and $Y$, respectively, and define the {\it entropic Ruzsa distance} by
$$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
This definition only depends on the individual distributions of $X$ and $Y$ and does not require them to
have the same sample space. When $X$ and $Y$ are independent, the Ruzsa distance enjoys some
nice properties.

@@

\proof By independence we have
$$2\dd\{X,Y\} = 2\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}\over 2,$$
Corollary~{\cormaxsumineq} tells us that $2\Eta\{X-Y\} \ge 2\Eta\{X\}$, so this is at least $\Eta\{X\}-\Eta\{Y\}$.
The same corollary also says that $2\Eta\{X-Y\} \ge 2\Eta\{Y\}$, which allows us to add the absolute value bars
to the inequality.

@@

@@

\proof Let $(X_1,Y_1)$ and $(X_2,Y_2)$ be conditionally independent trials of $(X,Y)$ relative
to $X-Y$. Since $(X,Y)$ determines $X-Y$, we have $X_1-Y_1 = =X-Y= X_2-Y_2$. Let $(X_3,Y_3)$ be another
trial of $(X,Y)$ independent of $(X_1, X_2, Y_1, Y_2)$. Then
$$X_3 + Y_3 = X_3 + Y_3 + X_1 - Y_1 - X_2 + Y_2 = (X_3-Y_2) - (X_1-Y_3) + X_2 + Y_1,$$
so $(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3, Y_3)$ each determine $X_3+ Y_3$. On the other hand,
$(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3,Y_3)$ together determine the sextuple $(X_1,X_2,X_3,Y_1,Y_2,Y_3)$,
so by the submodularity inequality, we have
\edef\eqsumdiff{\the\eqcount}
$$\eqalign{
\Eta\{X_1,X_2,X_3,Y_1,Y_2,&Y_3\} + \Eta\{X_3+Y_3\} \cr
&\le \Eta\{X_3-Y_2, X_1-Y_3, X_2, Y_1\} + \Eta\{X_3, Y_3\}.\cr
}\adveq$$
We have
$$\Eta\{X_3,Y_3\} = \Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$$
by independence of
$X$ and $Y$, and
since $(X_3,Y_3)$ and $(X_1,X_2,Y_1,Y_2)$ are independent, we have
$$\eqalign{
\Eta\{X_1,X_2,X_3,Y_1,Y_2,Y_3\} &= \Eta\{X_1,X_2,Y_1,Y_2\} + \Eta\{X_3,Y_3\} \cr
&= \Eta\{X_1, Y_1,X_2,Y_2,X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
&= 2\Eta\{X,Y,X-Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}\cr
&= 3\Eta\{X\} + 3\Eta\{Y\} - \Eta\{X-Y\},\cr
}$$
where in the third line we applied~\refeq{\eqcondindep}. On the other hand,
$$\Eta\{X_3 + Y_3\} = \Eta\{X+Y\}$$
and
$$\eqalign{
\Eta\{X_3-Y_2, &X_1-Y_3, X_2, Y_1\} \cr
&\le \Eta\{X_3-Y_2\} + \Eta\{X_1-Y_3\} + \Eta\{X_2\} + \Eta\{Y_1\} \cr
&= 2\Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}.
}$$
Substituting everything into~\refeq{\eqsumdiff} yields
$$\eqalign{
3\Eta\{X\} + 3\Eta\{Y\} - &\Eta\{X-Y\} + \Eta\{X+Y\} \cr
&\le 2\Eta\{X-Y\} + 2\Eta\{X\} + 2\Eta\{Y\},
}$$
and the desired inequality follows upon rearrangement of terms.\slug

@@

\proclaim Lemma \advthm. Let $X$, $Y$, and $Z$ be independent random variables taking values in a common
abelian group. Then
$$\Eta\{X+Y+Z\} - \Eta\{X+Y\} \le \Eta\{Y+Z\} -\Eta\{Y\}.$$

@@

@@

The fact that the implication goes the ``wrong'' way may seem somewhat baffling at first. We will now get
to the bottom of this.
If $A$ and $B$ are subsets of a finite abelian group, we can let $X$ and $Y$ be the uniform
distributions on $A$ and $B$, respectively. We have been operating under the belief that $\Eta\{X+Y\}$
should correspond (up to taking powers or logarithms) to the size of $A+B$. But this is not true, since
$X$ and $Y$ may be given a joint distribution that is not uniform on $A\times B$, even if
its marginals are uniform on $A$ and $B$.

@@

In the entropy setting, the operation on random variables that corresponds to taking subsets is
to conditioning. (As a sanity check, recall that conditioning never increases entropy, just
as taking subsets never increases cardinality.) The Balog--Szemer\'edi--Gowers theorem gives
us subsets between which we can take a {\it bona fide} sumset, so its entropic analogue
should return conditionings $X'$ and $Y'$ of $X$ and $Y$ relative to some random variable $Z$,
such that
\medskip
\item{i)} $X'$ and $Y'$ are conditionally independent relative to $Z$;
\smallskip
\item{ii)} the entropies $\Eta\{X'\given Z\}$ and
$\Eta\{Y'\given Z\}$ are not too small compared to their unconditioned analogues; and
\smallskip
\item{iii)} $\Eta\{X'+Y'\given Z\}$ is small.
\medskip
In fact, the conditioning we shall perform is exactly the one used to define additive energy.

@@

\edef\lematwo{\the\sectcount.\the\thmcount}
\proclaim Lemma \advthm. Let $X$ and $Y$ be discrete random variables taking values in the same abelian
group. Let $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
independent trials of $(X,Y)$ relative to $X+Y$. Then we have
$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} - \II\{X:Y\}.$$
The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\}$.

\proof First we perform the proof for $X_1 - Y_2$.
Submodularity gives us
$$\Eta\{X_1, Y_1, X_1-Y_2\} + \Eta\{X_1-Y_2\} \le
\Eta\{ X_1, X_1-Y_2\} + \Eta\{Y_1, X_1-Y_2\}.$$
Since $X_1 + Y_1 = X+Y = X_2 + Y_2$, given $(X_1, Y_1, X_1-Y_2)$ we can recover
the values of $X_2$ and $Y_2$. So $(X_1, Y_1, X_1-Y_2)$ and $(X_1,Y_1,X_2,Y_2)$ determine each
other and hence
$$\Eta\{X_1, Y_1, X_1-Y_2\} = \Eta\{X_1, Y_1, X_2, Y_2\} = 2\Eta\{X,Y\} - \Eta\{X+Y\}.$$
On the other side of the inequality, we have
$$\Eta\{ X_1, X_1-Y_2\} = \Eta\{X_1, Y_2\} \le \Eta\{X\} + \Eta\{Y\},$$
and similarly
$$\Eta\{ Y_1, X_1-Y_2\} = \Eta\{Y_1, X_2-Y_1\} = \Eta\{X_2, Y_1\} \le \Eta\{X\} + \Eta\{Y\}.$$
Therefore,
$$\eqalign{
\Eta\{X_1-Y_2\} &\le \Eta\{X+Y\} - 2\Eta\{X,Y\} + 2\Eta\{X\} + 2\Eta\{Y\} \cr
&= \Eta\{X+Y\} - \II\{X : Y\}.\cr
}$$
The same holds with the roles of $X_2$ and $Y_2$ exchanged.\slug

@@

But subspaces may not be a very efficient way of covering sets with small doubling. One can easily
cook up examples where there is enough linear independence in $A$ so that the smallest subspace
containing $A$ has size exponential in $|A|$.

Returning to first
principles, observe that if the doubling
constant of $A$ is $1$, then the structure of $A$ is completely determined, as shown by the following proposition.

@@

\advsect A compactness argument

@@

>>>>

\maketitle{Entropy and additive combinatorics}{}{Marcel K. Goh}{\sl Department of Mathematics and Statistics,
McGill University}

\floattext4.5 \ninebf Abstract.
\ninepoint These expository notes give a gentle introduction to the notion of entropy as it is used
in additive combinatorics, moving at a leisurely pace through the entropic analogues of Pl\"unnecke's
theorem and the Balog--Szemer\'edi--Gowers theorem before tackling the recent
proof of the polynomial Freiman--Ruzsa
conjecture by W.~T.~Gowers, B.~Green, F.~Manners, and T.~Tao. Effort has been put into making this
document as self-contained as possible, and extra proof details have been supplied in the hope
that these notes may be accessible to the average graduate student or enterprising undergraduate.

@@

We shall take it on faith that there really exists a function on random variables satisfying
these axioms.
(It is very possible you have met the formula for entropy
somewhere on your travels, but you will not find it anywhere in
these notes.)
In fact, the axioms only define entropy
up to a multiplicative constant, so we shall add the following axiom.

@@

Notationally, we would expect that $\Eta\{Y\given X\} = \Eta\{Y\}$ if $X$ and $Y$ are independent.
This is the first proposition we will carefully prove, using only the axioms. The axiomatic description above,
as well as many of the proofs in
this section, are transcribed (with some adaptations here and there) from lectures given by W.~T.~Gowers.

@@

Next we establish the intuitive statement
that the entropy of a uniform random variable supported on a set $A$ is at most
the entropy of a uniform random variable supported on a superset $B$ of $A$.

@@

The thread connecting entropy and additive combinatorics is rather a precarious one.
As noted by I.~Ruzsa~\ref{ruzsa2008}, there are scenarios in which combinatorial inequalities and
their corresponding entropy ones
are equivalent, scenarios in which they both hold but their equivalence cannot be established,
and further scenarios where an inequality holds in one world but cannot (or has not) been proven in the other.

The ``dictionary'' that allows us to translate
statements between cardinality inequalities and entropy inequalities is based, in part, on the following
observation. (In this and the rest of the notes, $\lg$ denotes the base-$2$ logarithm. This base can
be changed by modifying the normalisation axiom.)

@@

The submodularity inequality is often stated in terms of a triple of random variables in terms of
the {\it conditional mutual information}, which is defined by
$$\eqalign{
\II\{ X : Y \given Z\} &= \sum_{z\in C} p_Z(z) \II\bigl\{ (X\given Z = z) : (Y\given Z = z)\bigr\} \cr
&= \Eta\{X\given Z\} - \Eta\{X\given Y, Z\} \cr
&= \Eta\{Y\given Z\} - \Eta\{Y\given X, Z\}.\cr
}$$

@@

\proof Since conditioning does not increase entropy, we have
$$\Eta\{ X\pm Y\} \ge \Eta\{ X\pm Y \given Y\},$$
and since the probabilities $\pr\{ X + Y = z \given Y = y\} = \pr\{ X = z-y \given Y = y\}$ for all $z\in G$,
by invariance we have
$$\Eta\{ X\pm Y\} \ge \Eta\{X\given Y\} = \Eta\{X\} - \II\{X : Y\}.$$
Repeating the same argument but exchanging the r\^oles of $X$ and $Y$, we get
$$\Eta\{ X\pm Y\} \ge \Eta\{Y\given X\} = \Eta\{Y\} - \II\{X : Y\},$$
so
$$\Eta\{ X\pm Y\} \ge \max\bigl(\Eta\{X\}, \Eta\{Y\}\bigr) - \II\{X : Y\}.$$
Now let $Z$ be any random variable with finite support.
$$\eqalign{
\Eta\{ X\pm Y\given Z\} &= \sum_{z\in G} \pr\{Z = z\} \Eta\{ X \pm Y \given Z = z\} \cr
&\ge \Bigl(\max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} \Bigr)
\sum_{z\in G} \pr\{Z = z\} \cr
&= \max\bigl(\Eta\{X\given Z\}, \Eta\{Y\given Z\}\bigr) - \II\{X : Y\given Z\} ,\cr
}$$
which completes the proof.\slug

@@

\medskip\boldlabel Entropic Ruzsa distance.
In additive combinatorics, whenever we have two finite subsets $A$ and $B$
of the same abelian group, we can compute the Ruzsa distance
$$d(A,B) = \lg {|A-B|\over \sqrt{|A|\cdot|B|}}$$
between them. (This satisfies all the axioms of a metric except the one requiring $d(A,A) = 0$ for all sets
$A$.) This ``distance'' is called the {\it Ruzsa distance} as it was first defined by I.~Ruzsa~\ref{ruzsa1996}.

The entropic analogue of the Ruzsa distance is defined as follows.
For finitely supported random variables $X$ and $Y$ taking values in the same abelian group,
we let $X'$ and $Y'$ be independent copies of $X$
and $Y$, respectively, and define the {\it entropic Ruzsa distance} by
$$\dd\{X,Y\} = \Eta\{X'-Y'\} - {\Eta\{X'\} \over 2} - {\Eta\{Y'\}\over 2}.$$
This definition, first established by T.~Tao~\ref{tao2010},
only depends on the individual distributions of $X$ and $Y$ and does not require them to
have the same sample space. When $X$ and $Y$ are independent, the entropic Ruzsa distance enjoys some
nice properties.

@@

\proof By independence we have
$$2\dd\{X,Y\} = 2\Eta\{X-Y\} - \Eta\{X\} - \Eta\{Y\}.$$
Corollary~{\cormaxsumineq} tells us that $2\Eta\{X-Y\} \ge 2\Eta\{X\}$, so this is at least $\Eta\{X\}-\Eta\{Y\}$.
The same corollary also says that $2\Eta\{X-Y\} \ge 2\Eta\{Y\}$, which allows us to add the absolute value bars
to the inequality.

@@

The following inequality relates the conditional and unconditional versions of Ruzsa distance.

\edef\lemfiveone{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma 5.1). Let $(X,Z)$, and $(Y,W)$ be random
variables, with $X$ and $Y$ taking values in the same abelian group. Then
$$\dd\{ X\given Z ; Y\given W\} \le \dd\{X,Y\} + {1\over 2} \II\{X:Z\} + {1\over 2}\II\{Y:W\}.$$

\proof Letting $(X',Z')$ and $(Y',W')$ be independent copies of the given random variables, we
expand
$$\eqalign{
\dd\{ X\given Z ; Y\given W\}
&= \Eta\{X'-Y'\given Z',W'\} - {\Eta\{X'\given Z'\}\over 2} - {\Eta\{Y'\given W'\}\over 2} \cr
&\le \Eta\{X'-Y'\} - {\Eta\{X',Z'\}-\Eta\{Z'\}\over 2} \cr
&\hskip130pt- {\Eta\{Y', W'\}-\Eta\{W'\}\over 2} \cr
&= \dd\{X-Y\} + {\Eta\{X'\} + \Eta\{Z'\} -\Eta\{X',Z'\}\over 2} \cr
&\hskip90pt+ {\Eta\{Y'\} + \Eta\{W'\} - \Eta\{Y', W'\}\over 2} \cr
&= \dd\{X-Y\} + {1\over 2}\II\{X:Z\} + {1\over 2}\II\{Y:W\},\cr
}$$
by independence and the fact that conditioning does not increase entropy.\slug

@@

\proof Let $(X_1,Y_1)$ and $(X_2,Y_2)$ be conditionally independent trials of $(X,Y)$ relative
to $X-Y$. Since $(X,Y)$ determines $X-Y$, we have $X_1-Y_1 =X-Y= X_2-Y_2$. Let $(X_3,Y_3)$ be another
trial of $(X,Y)$ independent of $(X_1, X_2, Y_1, Y_2)$. Then
$$X_3 + Y_3 = X_3 + Y_3 + X_1 - Y_1 - X_2 + Y_2 = (X_3-Y_2) - (X_1-Y_3) + X_2 + Y_1,$$
so $(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3, Y_3)$ each determine $X_3+ Y_3$. On the other hand,
$(X_3-Y_2, X_1-Y_3, X_2, Y_1)$ and $(X_3,Y_3)$ together determine the sextuple $(X_1,X_2,X_3,Y_1,Y_2,Y_3)$,
so by the submodularity inequality, we have
\edef\eqsumdiff{\the\eqcount}
$$\eqalign{
\Eta\{X_1,X_2,X_3,Y_1,Y_2,&Y_3\} + \Eta\{X_3+Y_3\} \cr
&\le \Eta\{X_3-Y_2, X_1-Y_3, X_2, Y_1\} + \Eta\{X_3, Y_3\}.\cr
}\adveq$$
We have
$$\Eta\{X_3,Y_3\} = \Eta\{X,Y\} = \Eta\{X\} + \Eta\{Y\}$$
by independence of
$X$ and $Y$, and
since $(X_3,Y_3)$ and $(X_1,X_2,Y_1,Y_2)$ are independent, we have
$$\eqalign{
\Eta\{X_1,X_2,X_3,Y_1,Y_2,Y_3\} &= \Eta\{X_1,X_2,Y_1,Y_2\} + \Eta\{X_3,Y_3\} \cr
&= \Eta\{X_1, Y_1,X_2,Y_2,X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
&= 2\Eta\{X,Y,X-Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}\cr
&= 3\Eta\{X\} + 3\Eta\{Y\} - \Eta\{X-Y\},\cr
}$$
where in the third line we applied~\refeq{\eqcondindep}. On the other hand,
$$\Eta\{X_3 + Y_3\} = \Eta\{X+Y\}$$
and
$$\eqalign{
\Eta\{X_3-Y_2, &X_1-Y_3, X_2, Y_1\} \cr
&\le \Eta\{X_3-Y_2\} + \Eta\{X_1-Y_3\} + \Eta\{X_2\} + \Eta\{Y_1\} \cr
&= 2\Eta\{X-Y\} + \Eta\{X\} + \Eta\{Y\}.
}$$
Substituting everything into~\refeq{\eqsumdiff} yields
$$\eqalign{
3\Eta\{X\} + 3\Eta\{Y\} - &\Eta\{X-Y\} + \Eta\{X+Y\} \cr
&\le 2\Eta\{X-Y\} + 2\Eta\{X\} + 2\Eta\{Y\},
}$$
and the desired inequality follows upon rearrangement of terms.\slug

@@

\edef\lemaone{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma A.1).
Let $X$, $Y$, and $Z$ be independent random variables taking values in a common
abelian group. Then
$$\Eta\{X+Y+Z\} - \Eta\{X+Y\} \le \Eta\{Y+Z\} -\Eta\{Y\}.$$

@@

We conclude this section by recording two consequences of Lemma~{\lemaone} and will
be of use in the proof of the polynomial Freiman--Ruzsa theorem.

\edef\lemfivetwo{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma 5.2).
Let $X$, $Y$, and $Z$ be random variables taking values in an abelian group,
with $Y$ and $Z$ independent. Then
$$\eqalign{
\dd\{X, Y-Z\} - \dd\{X, Y\} &\le {\Eta\{Y-Z\} - \Eta\{Y\}\over 2} \cr
&= {1\over 2} \dd\{Y, Z\} + {1\over 4}\Eta\{Z\} - {1\over 4}\Eta\{Y\}\cr
}$$
and
$$\eqalign{
\dd\{X; Y\given Y-Z\} - \dd\{X, Y\} &\le {\Eta\{Y-Z\} - \Eta\{Z\}\over 2} \cr
&= {1\over 2} \dd\{Y, Z\} + {1\over 4}\Eta\{Z\} - {1\over 4}\Eta\{Z\}.\cr
}$$

\proof For the first inequality, let $X'$ be a copy of $X$ that is independent of $(Y,Z)$, so that
$$\eqalign{
\dd\{X, Y-Z\} - \dd\{X, Y\}
&= \Eta\{X'-Y+Z\} - {1\over 2}\Eta\{X'\} - {1\over 2} \Eta\{Y-Z\} \cr
&\hskip60pt-\Eta\{X'-Y\} + {1\over 2}\Eta\{X'\} + {1\over 2}\Eta\{Y\}. \cr
&= \Eta\{X'-Y+Z\} - {1\over 2} \Eta\{Y-Z\} \cr
&\hskip110pt-\Eta\{X'-Y\} + {1\over 2}\Eta\{Y\}.\cr
}$$
Lemma~\ref{\lemaone} with $X'$ in place of $X$ and $-Y$ in place of $Y$ yields
$$\Eta\{X'-Y+Z\} - \Eta\{X'-Y\} \le \Eta\{Y-Z\} -\Eta\{Y\},$$
so
$$\dd\{X, Y-Z\} - \dd\{X, Y\} \le {1\over 2} \Eta\{Y-Z\} - {1\over 2}\Eta\{Y\},$$
as desired. The alternate version of the statement follows directly from the definition of Ruzsa distance,
since $Y$ and $Z$ are independent.

For the other inequality, we apply Lemma~\ref{\lemfiveone} with $W = Y-Z$ (and the variable $Z$ in that lemma
set to anything that is independent of $X$) to obtain
$$\eqalign{
\dd\{X; Y\given Y-Z\} - \dd\{X,Y\} &\le {1\over 2} \II\{Y:Y-Z\} \cr
&= {1\over 2}\bigl(\Eta\{Y\} + \Eta\{Y-Z\} - \Eta\{Y, Y-Z\}\bigr) \cr
&= {1\over 2}\bigl(\Eta\{Y\} + \Eta\{Y-Z\} - \Eta\{Y, Z\}\bigr) \cr
&= {1\over 2}\bigl(\Eta\{Y-Z\} - \Eta\{Z\}\bigr),\cr
}$$
where in the last line we used independence of $Y$ and $Z$. Independence also gives the alternative version
of the right-hand-side expression.\slug

Changing variables in the first inequality and then adding the second inequality gives us the following.

\edef\lemsevenone{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma 7.1).
Let $X$, $Y$, $Z$, and $W$ be random variables taking values in the same abelian group, with $Y$, $Z$,
and $W$ independent. Then
$$\eqalign{
\dd\{&X; Y-Z\given Y-Z-W\} - \dd\{X,Y\} \cr
&\le {1\over 2} \bigl( \Eta\{Y-Z-W\} + \Eta\{Y-Z\} - \Eta\{Y\} - \Eta\{W\}\bigr) .\cr
}$$

\proof The second inequality of Lemma~\ref{\lemfivetwo} (setting $Y \gets Y-Z$ and $Z\gets W$) yields
$$\dd\{X; Y-Z\given Y-Z-W\} - \dd\{X, Y-Z\} \le {\Eta\{Y-Z-W\} - \Eta\{W\}\over 2},$$
and the first inequality (without any change of variables) is
$$\dd\{X, Y-Z\} - \dd\{X, Y\} \le {\Eta\{Y-Z\} - \Eta\{Y\}\over 2}.$$
Adding these two inequalites proves the lemma.\slug

@@

The fact that the implication goes the ``wrong'' way may seem somewhat baffling. Let us get
to the bottom of this.
If $A$ and $B$ are subsets of a finite abelian group, we can let $X$ and $Y$ be the uniform
distributions on $A$ and $B$, respectively. We have been operating under the belief that $\Eta\{X+Y\}$
should correspond (up to taking powers or logarithms) to the size of $A+B$. But this is not true, since
$X$ and $Y$ may be given a joint distribution that is not uniform on $A\times B$, even if
its marginals are uniform on $A$ and $B$.

@@

In the entropy setting, the operation on random variables that corresponds to taking subsets is
conditioning. (As a sanity check, recall that conditioning never increases entropy, just
as taking subsets never increases cardinality.) The Balog--Szemer\'edi--Gowers theorem gives
us subsets between which we can take a {\it bona fide} sumset, so its entropic analogue
should return conditionings $X'$ and $Y'$ of $X$ and $Y$ relative to some random variable $Z$,
such that
\medskip
\item{i)} $X'$ and $Y'$ are conditionally independent relative to $Z$;
\smallskip
\item{ii)} the entropies $\Eta\{X'\given Z\}$ and
$\Eta\{Y'\given Z\}$ are not too small compared to their unconditioned analogues; and
\smallskip
\item{iii)} $\Eta\{X'+Y'\given Z\}$ is small.
\medskip
In fact, the conditioning we shall perform is exactly the one used to define additive energy.

@@

\edef\lematwo{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma A.2).
Let $X$ and $Y$ be discrete random variables taking values in the same abelian
group. Let $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
independent trials of $(X,Y)$ relative to $X+Y$. Then we have
$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} - \II\{X:Y\}.$$
The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\}$.

\proof First we perform the proof for $X_1 - Y_2$.
Submodularity gives us
$$\Eta\{X_1, Y_1, X_1-Y_2\} + \Eta\{X_1-Y_2\} \le
\Eta\{ X_1, X_1-Y_2\} + \Eta\{Y_1, X_1-Y_2\}.$$
Since $X_1 + Y_1 = X+Y = X_2 + Y_2$, given $(X_1, Y_1, X_1-Y_2)$ we can recover
the values of $X_2$ and $Y_2$. So $(X_1, Y_1, X_1-Y_2)$ and $(X_1,Y_1,X_2,Y_2)$ determine each
other and hence
$$\Eta\{X_1, Y_1, X_1-Y_2\} = \Eta\{X_1, Y_1, X_2, Y_2\} = 2\Eta\{X,Y\} - \Eta\{X+Y\}.$$
On the other side of the inequality, we have
$$\Eta\{ X_1, X_1-Y_2\} = \Eta\{X_1, Y_2\} \le \Eta\{X\} + \Eta\{Y\},$$
and similarly
$$\Eta\{ Y_1, X_1-Y_2\} = \Eta\{Y_1, X_2-Y_1\} = \Eta\{X_2, Y_1\} \le \Eta\{X\} + \Eta\{Y\}.$$
Therefore,
$$\eqalign{
\Eta\{X_1-Y_2\} &\le \Eta\{X+Y\} - 2\Eta\{X,Y\} + 2\Eta\{X\} + 2\Eta\{Y\} \cr
&= \Eta\{X+Y\} - \II\{X : Y\}.\cr
}$$
The same holds with $Y_2$ replaced by $X_2$.\slug

@@

But subspaces may not be a very efficient way of covering sets with small doubling. One can easily
cook up examples where there is enough linear independence in $A$ so that the smallest subspace
containing $A$ has size exponential in $|A|$, but $A$ still has small doubling.

On the other hand, in the extreme case that the doubling
constant of $A$ is $1$, then the structure of $A$ is completely determined, as shown by the following proposition.

@@

\edef\sectcompactness{\the\sectcount}
\advsect A compactness argument

@@

\advsect Entropy distance under homomorphisms

Our goal has been reduced to showing that if $X_1$ and $X_2$ have $\dd\{X_1, X_2\} > 0$, then
there exist $X_1'$ and $X_2'$ with $\tau\{X_1', X_2'\} < \tau\{X_1, X_2\}$. Remember that
the fixed variables $X_1^*$ and $X_2^*$ still play a r\^ole in this statement, since
they feature in the definition of the functional $\tau$.

We have elected to first prove all the technical lemmas
then stitch them into a complete proof.
We employ this ``bottom-up'' approach not because
it is necessarily more natural, but because the original proof is written
in a rather ``top-down'' style, and the reader may find it enlightening to have both expositions at his
or her disposal.

Having said this, the remainder of this section will be devoted to the study of how entropy
behaves under homomorphism. This appeared as Proposition~1.4
in~\ref{revisited2023}, and reproved in~\ref{pfr2023} with the
explicit error term shown below.

\edef\propfourone{\the\sectcount.\the\thmcount}
\parenproclaim Proposition {\advthm} ({\rm\ref{pfr2023},} Proposition 4.1).
Let $\pi : G\to G'$ be a homomorphism of abelian groups and let
$Z_1$ and $Z_2$ be $G$-valued random variables. Then
$$\dd\{Z_1, Z_2\} \ge \dd\bigl\{ \pi(Z_1), \pi(Z_2)\bigr\}
+ \dd\bigl\{ Z_1\given \pi(Z_1); Z_2\given \pi(Z_2)\bigr\}.$$
If $Z_1$ and $Z_2$ are assumed to be independent, then the two sides differ by
$$\II\bigl\{ Z_1 - Z_2 : \bigl( \pi(Z_1), \pi(Z_2)\bigr) \mathbin{\big|} \pi(Z_1-Z_2)\bigr\}.$$

\proof Let $Z_1'$ and $Z_2'$ be independent copies of $Z_1$ and $Z_2$. Then
$$\eqalign{
\dd\bigl\{ Z_1\given &\pi(Z_1); Z_2\given \pi(Z_2)\bigr\} \cr
&= \Eta\bigl\{Z_1'-Z_2'\given \pi(Z_1'), \pi(Z_2') \bigr\}
 - {1\over 2}\Eta\bigl\{ Z_1' \given \pi(Z_1')\bigr\} - {1\over 2}\Eta\bigl\{ Z_2' \given \pi(Z_2')\bigr\} \cr
&\le \Eta\bigl\{Z_1'-Z_2'\given \pi(Z_2') \bigr\}
 - {1\over 2}\Eta\bigl\{ Z_1' \given \pi(Z_1')\bigr\} - {1\over 2}\Eta\bigl\{ Z_2' \given \pi(Z_2')\bigr\} \cr
&= \Eta\bigl\{Z_1'-Z_2'\given \pi(Z_1'- Z_2') \bigr\}
 - {1\over 2}\Eta\bigl\{ Z_1' \given \pi(Z_1')\bigr\} - {1\over 2}\Eta\bigl\{ Z_2' \given \pi(Z_2')\bigr\} \cr
&= \Eta\{Z_1'-Z_2'\} - \Eta\bigl\{ \pi(Z_1'-Z_2')\bigr\} \cr
&\qquad\qquad
+{1\over 2}\Eta\{Z_1'\} - {1\over 2}\Eta\bigl\{ \pi(Z_1')\bigr\} 
+{1\over 2}\Eta\{Z_2'\} - {1\over 2}\Eta\bigl\{ \pi(Z_2')\bigr\} \cr
&= \dd\{Z_1, Z_2\} - \dd\bigl\{ \pi(Z_1),\pi(Z_2)\bigr\},\cr
}$$
where the inequality follows from submodularity, and in the second-last equality we used (three times)
the identity
$$\Eta\bigl\{X\given \pi(X)\bigr\} = \Eta\{X\} - \Eta\bigl\{\pi(X)\bigr\},$$
which holds for any $G$-valued random variable $X$, since $X$ determines $\pi(X)$.
